Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-06-11 paper.qmd
Added the notation table to appendix

- Remove all the extraneous commented text (NOT Done)
- Check how \kD{} and \gD{} are used, and make sure it's sensible
- Check spelling is consistently American
- Check references
- Add videos - make a list of the videos in the supplementary section, with the number of the figure that it corresponds too. This avoids having captions that are too long and awkward with raw URLs. And it keeps all of this supplementary material in one place.
- Move the notation table to appendix
- Code clean up: recommend that all of the calculations done with quollr are done with a separate script(s) and put in a scripts directory. Save the resulting data need to make plots and tables. Only code for the plots and tables is directly in the qmd in the final version.

@@

```{r}
#| label: tbl-notation
#| tbl-cap: "Summary of notation for describing new methodology."
# Notation used in the paper

notation_df <- read_csv("misc/notation.csv")

# Create the table
kable(notation_df, 
      format = "latex", 
      booktabs = TRUE, escape = FALSE) |>
  kable_styling(position = "center", 
                full_width = FALSE, 
                font_size = 12) |>
  row_spec(0, bold = TRUE) |>
  column_spec(1:2, width = c("3cm", "12cm"))
```

We define the NLDR as a function $g\text{:}~ \mathbb{R}^{n\times p} \rightarrow \mathbb{R}^{n\times k}$, with (hyper-)parameters $\mathbfit{\theta}$. The parameters, $\mathbfit{\theta}$, depend on the choice of $g$, and can be considered part of model fitting in the traditional sense. Common choices for $g$ include functions used in tSNE, UMAP, PHATE, TriMAP, PaCMAP, or MDS, although in theory any function that does this mapping is suitable. <!--Any input requirements for the data (such as normalization, or preprocessing through the use of PCA or similar) is considered part of the function $g$.-->

@@

# error_plot_two_curvy_hist <- 
#   ggplot(error_df_two_curvy_abs) +
#   geom_histogram(aes(x=sqrt_row_wise_total_error, y=..density..)) +
#   geom_density(aes(x=sqrt_row_wise_total_error, y=..density..), colour="red") +
#   #xlab(expression(group("|", e[hj], "|"))) +
#   xlab(expression(e[hj])) +
#   ylab("") +
#   theme_bw()

@@

two_curvy_proj_tsne_all_model1 <- plot_proj(
  proj_obj = proj_obj2, 
  point_param = c(0.1, 0.5, clr_choice), # size, alpha, color
  line_param = c(0.5, 0.5, "black"), # linewidth, alpha
  plot_limits = c(-0.5, 0.5), 
  axis_text_size = 2, 
  is_category = FALSE) +
  interior_annotation(label = "c", position = sc_ltr_pos, cex = 1.2) +
  theme(aspect.ratio = 1) #+
  #geom_segment(
  #  data = model_df, 
  #  aes(
  #    x = proj1_from, 
  #    y = proj2_from, 
  #    xend = proj1_to, 
  #    yend = proj2_to), 
  #  color = "#000000")

@@

<!-- XXX How are you doing this? Do you check the bounds of the hexagon? Or do you use distance to centroid? -->

@@

Observations are associated with their bin center, $C_{h}^{(p)}$, which are also considered to be the *fitted values*. These can also be denoted as $\widehat{X}$. <!--The fitted values of the points in $H_h$ refers to the \pD{} mapping $C_{h}^{(p)}$ of the corresponding \kD{} model point $C_{h}^{(2)}$.-->

<!--#### Error-->

@@

# df_bin_centroids_gau1 <- df_bin_centroids_gau1 |>
#   dplyr::filter(std_counts > 0)

@@

trimesh_removed_gau_tsne + trimesh_removed_gau_umap + trimesh_removed_gau_pacmap +
five_gau_proj_tsne_model1 + five_gau_proj_umap_model1 + five_gau_proj_pacmap_model1 + 
  #five_gau_proj_tsne_model2 +
  #five_gau_proj_umap_model2 + five_gau_proj_pacmap_model2 +
  plot_layout(guides = "collect", nrow = 2) &
  theme(legend.position='none')
```

@@

<!-- To showcase the robustness of our method when applied to data with different characteristics, this section presents two case studies. The first application focuses on single-cell data, a domain where the high dimensionality, sparsity, and heterogeneity of gene expression measurements pose significant analytical challenges. Our approach facilitates the identification of the most accurate NLDR layout, enabling the discovery of cell populations with similar expression profiles. The second case study explores the MNIST hand-written digits dataset, which serves as a benchmark for evaluating the preservation of local structures, specifically with digit 1.-->

@@

This is a benchmark single-cell data set collected on Human Peripheral Blood Mononuclear Cells (PBMC3k) as used in @pbmc2019. Single-cell data measures the gene expression of individual cells in a sample of tissue (see  for example, @haque2017). This type of data is used to obtain an understanding of cellular level behavior and heterogeneity in their activity. Clustering of single-cell data is used to identify groups of cells with similar expression profiles. NLDR often used to summarise the cluster structure. Usually, NLDR does not use the cluster labels to compute the layout, but uses color to represent the cluster labels when it is plotted. <!-- The purpose of this example is to *illustrate how to use our method to help decide on an appropriate NLDR layout that accurately represents the data*. The cluster results of Human Peripheral Blood Mononuclear Cells (PBMC3k) in @chen2024 are examined.-->

@@

<!-- The @fig-NLDR-variety (a) shows three well-separated clusters with big separations. However, as shown in @fig-model-pbmc-author-proj (a2), there is no big separation between three clusters in $9\text{-}D$. Therefore, the suggested UMAP representation (@fig-NLDR-variety (a)) does not accurately represent the structure of PBMC3k dataset.-->

@@

<!-- Layouts c, d, and e, which show small separations between clusters, are universally optimal. However, layout d performs well with smaller binwidths and poorly with larger binwidths. On the other hand, layout e performs well with larger binwidths. Layout c was selected for further analysis due to its better (hyper-)parameter selection for the same method of the published plot.-->

<!-- XXX can we use a and e in these figures instead of a and b, so it is easier to describe relative to the first figure, and previous figure? -->

@@

<!-- The visualization of the selected layout in the $9\text{-}D$ shows edges between the clusters (@fig-model-pbmc-author-proj (b2)). This supports the presence of small separations between clusters. Additionally, the data points are not uniformly distributed across the clusters, resulting in dense areas (@fig-model-pbmc-author-proj (b2)). Furthermore, the clusters shows non-linear shapes (@fig-model-pbmc-author-proj (b3)).-->

@@

The digit "1" of the MNIST dataset [@lecun2010] consists of $7877$ grayscale images of handwritten "1"s. Each image is $28 \times 28$ pixels which corresponds to 784 variables. The first 10 principal components, explaining 83% of the total variation, are used. This data is essentially lies on a nonlinear manifold in the high dimensions, defined by the shapes that "1"s make in when sketched, menaing that some pixels are almost always dark and some almost always light. We expect that the best layout captures this type of structure and does not exhibit distinct clusters. <!-- The objective is to select a reasonable \gD{} layout, representing the non-linear structure of the digit 1 dataset in $10\text{-}D$ (@fig-clust-mnist (a)). -->

@@

<!-- The MSE for different binwidths ($a_1$) using UMAP, PHATE, PaCMAP, TriMAP with default (hyper-)parameter setting and tSNE with default (hyper-)parameter setting and perplexity $89$ (@fig-mnist-mse) were calculated. It is found that tSNE with perplexity $89$ (@fig-mnist-mse (f)) provide the most reasonable representation for the digit 1 dataset, showing universally best. However, \gD{} representation shows a big non-linear cluster and a small cluster with a small gap.-->

<!-- In the case of digit 1 data, there should not be any clusters unless anomalies exist, indicating different digit 1 patterns. The angle of the digit 1 images varies along this non-linear clustering structure (@fig-model-error-mnist), while the small cluster contains the digit 1 images with different patterns of the digit 1, unlike the usual (@fig-model-error-mnist (c)). This provides the evidence for two close clusters.

By visualizing the model generated for tSNE in $10\text{-}D$ helps to assess the \gD{} layout. As shown in @fig-clust-mnist (a1), the model provides the evidence for the non-linear structure of the digit 1 data in $10\text{-}D$. The model shows some quirks. The model's twisted pattern provides evidence for the $10\text{-}D$ data structure, which is not observed by \gD{} layout (@fig-clust-mnist (a2)). -->

>>>>

- Remove all the extraneous commented text (Done)
- Check how \kD{} and \gD{} are used, and make sure it's sensible
- Check spelling is consistently American
- Check references
- Add videos - make a list of the videos in the supplementary section, with the number of the figure that it corresponds too. This avoids having captions that are too long and awkward with raw URLs. And it keeps all of this supplementary material in one place.
- Move the notation table to appendix
- Code clean up: recommend that all of the calculations done with quollr are done with a separate script(s) and put in a scripts directory. Save the resulting data need to make plots and tables. Only code for the plots and tables is directly in the qmd in the final version.

@@

We define the NLDR as a function $g\text{:}~ \mathbb{R}^{n\times p} \rightarrow \mathbb{R}^{n\times k}$, with (hyper-)parameters $\mathbfit{\theta}$. The parameters, $\mathbfit{\theta}$, depend on the choice of $g$, and can be considered part of model fitting in the traditional sense. Common choices for $g$ include functions used in tSNE, UMAP, PHATE, TriMAP, PaCMAP, or MDS, although in theory any function that does this mapping is suitable.

@@

@@

two_curvy_proj_tsne_all_model1 <- plot_proj(
  proj_obj = proj_obj2, 
  point_param = c(0.1, 0.5, clr_choice), # size, alpha, color
  line_param = c(0.5, 0.5, "black"), # linewidth, alpha
  plot_limits = c(-0.5, 0.5), 
  axis_text_size = 2, 
  is_category = FALSE) +
  interior_annotation(label = "c", position = sc_ltr_pos, cex = 1.2) +
  theme(aspect.ratio = 1)

@@

@@

Observations are associated with their bin center, $C_{h}^{(p)}$, which are also considered to be the *fitted values*. These can also be denoted as $\widehat{X}$. 
<!--#### Error-->

@@

@@

trimesh_removed_gau_tsne + trimesh_removed_gau_umap + trimesh_removed_gau_pacmap +
five_gau_proj_tsne_model1 + five_gau_proj_umap_model1 + five_gau_proj_pacmap_model1 + 
  plot_layout(guides = "collect", nrow = 2) &
  theme(legend.position='none')
```

@@

@@

This is a benchmark single-cell data set collected on Human Peripheral Blood Mononuclear Cells (PBMC3k) as used in @pbmc2019. Single-cell data measures the gene expression of individual cells in a sample of tissue (see  for example, @haque2017). This type of data is used to obtain an understanding of cellular level behavior and heterogeneity in their activity. Clustering of single-cell data is used to identify groups of cells with similar expression profiles. NLDR often used to summarise the cluster structure. Usually, NLDR does not use the cluster labels to compute the layout, but uses color to represent the cluster labels when it is plotted.

@@

@@

@@

@@

The digit "1" of the MNIST dataset [@lecun2010] consists of $7877$ grayscale images of handwritten "1"s. Each image is $28 \times 28$ pixels which corresponds to 784 variables. The first 10 principal components, explaining 83% of the total variation, are used. This data is essentially lies on a nonlinear manifold in the high dimensions, defined by the shapes that "1"s make in when sketched, menaing that some pixels are almost always dark and some almost always light. We expect that the best layout captures this type of structure and does not exhibit distinct clusters.

@@
